R Notebook

### 2017-03-30 manuscript/draft.Rmd

@@

### Models

**Baseline models**
* The lme4 model fitted above.  This model assumes that the site-level average
  is constant, and tries to estimate its value.  The confidence intervals don't
  expand over time (except when observers change) because the model says nothing
  has changed.
* Naive AR1 model (fitted to individual sites' residuals from the observer model). 
  This model says that richness can diffuse up or down from the previous year's 
  value, without any constraints. As a result, the confidence intervals expand 
  quickly over time.
* Fit once per MCMC sample, to account for uncertainty in what observer effect
  values we should subtract off

**Time series models**
* These models only looked at one site at a time and looked for patterns in 
  univariate time series.
* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
  model selection based on AICc.
    * Autoregressive: current state depends on previous state(s)
    * Moving average: current observation error depends on previous observation error
    * Integrated/differencing: is analysis performed on raw data, on sequential 
      differences of the raw data, on sequential differences of sequential differences,
      etc.
    * default settings, except seasonal = FALSE
* Auto-ARIMA + site-level environmental variables over time.
* Uncertainty treated as above

**"Macroecological" model**
* Use a nonlinear model to predict richness as a function of environmental predictors
  (BIOCLIM + elevation + NDVI).
* We chose the `gbm` package for fitting boosted regression trees as our example of
  a nonlinear "macroecological model".
    * Works by fitting thousands of small models sequentially, and then adding up
      their predictions
    * the individual small models were regression trees with [[settings]].
* Compared to SDMs, this lets us predict richness directly instead of using hundreds of
  noisy species-level estimates, but doesn't use the full data set in one model.
* For this part of the analysis, we used the `gbm` package for fitting boosted regression
  trees with Gaussian errors.  
* We included observer effects & the associated uncertainty by running this model [[N]]
  times (once per MCMC sample).
* We did *not* include the site-level random effects as predictors, meaning that this
  approach implicitly assumes that two sites with identical Bioclim, elevation, and 
  NDVI values should have identical richness distributions.

**SDMs**
* Probably the most common approach?
* These models say that species' occurrence probabilities depend only on the values of
  the environmental variables at the time and place of the transect, not on the history
  of the environment or the history of the birds at that site
* Predict each species' occurrence probabilities individually, then add up the predictions
* [[Something about not using thresholds that throw away information]]
* Taking this approach in the face of uncertainty about observer effects would either 
  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
  times the number of species.
* To make it feasible to fit this large number of models, we used a custom wrapper 
  around the randomForest package. Random forests are flexible models that are built 
  by combining many independent models ("trees") fit to different versions of the training
  data, which made it straightforward to adapt them to our situation: we simply fit one 
  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
  predictions of the resulting forest.
* For the uncertainty, we used Gaussian approximations, based on the means and variances
  of the sums of independent Bernoulli random variables (cite Calabrese et al.).

@@

**Ensembles**
* Ensembles. Averaging predictions from multiple models tends to reduce the 
  noise in the estimates, and can thus lead to better predictions. Choosing the
  weights for the ensemble isn't straightforward because our estimates of model 
  error on the training set are biased in favor of models that
  overfit. [[Another option is to spatially cross-validate on the training set]]

**Evaluation**

* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
* Metrics:
  * (R)MSE 
    * How far, on average, are the models' predictions from the true value?
    * [[RMSE has better units than MSE (species versus squared species), but MSE is
      nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
  * Calibration/bias
    * In general, or in specific situations, is the model consistently predicting
      that richness will be too high or too low?
    * Can also report this value in terms of RMSE or MSE
  * Coverage of 95% confidence intervals
  * Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
    and coverage)

>>>>

`r settings = yaml::yaml.load_file("settings.yaml")`

@@

### Site-level models

[[note that these all use a Gaussian error distribution]]

**Baseline models.** The two simplest models we fit treated site-level richness
observations either as uncorrelated noise around a site-level constant (the 
"average" model) or as an autoregressive model with a single year of history 
(the "naive" model). The predictions from these two models are qualitatively 
different: predictions from the "average" model are centered on the average 
richness and the confidence intervals are narrow and constant-width. The
"naive" model, in contrast, predicts that future observations will be similar to
the final observed value (e.g. the value observed in `$r settings$last_train_year`), 
and the confidence intervals expand rapidly as the predictions extend farther into 
the future.  Each of these model types was fit separately to each BBS route in two
ways: first, we fit the model to the raw richness values, without accounting for
possible differences among observers at the site.  Then, we fit the models to the
residuals from our observer model (i.e. fitting the model to an estimate of how
many species a typical observer would have found, rather than to the raw data). 
In the latter case, we averaged across [[N]] models that were fit separately to 
the [[N]] Monte Carlo estimates of the observer effects, to account for our
uncertainty in the true values of those effects.

**Time series models.** Like the two baseline models, these models only looked at 
one site at a time, but they could include additional complexity. As above, these
models were trained both with the raw richness time series and with Monte Carlo
estimates of the observer-corrected time series. The  Auto-ARIMA models (based on 
Hyndman's 2016 `auto.arima` function) could include an autoregressive component 
(as in the "naive" model, but with the possibility of longer-term dependencies
in the underlying process), a moving average component (where the noise can have 
serial autocorrelation) and an integration/differencing component (so that the 
analysis could be performed on sequential differences of the raw data, rather than
on the observed time series itself). The `auto.arima` function chooses whether to 
include each of these components, and how many terms to include for each one, using
AICc, as described in Hyndman (2016). We used the default settings for this function,
apart from setting `seasonal = FALSE` because our time series does not include a 
seasonal component.  [[Note: should probably record somewhere what order model
the auto.arima function tends to choose.  Eyeballing the results, seems to be most 
similar to the naive model, but would be good to check.]] We also fit an 
"Auto-ARIMA + environment" model, in which each site-level time series could 
also be predicted by annual changes in our environmental variables at that site.

### Continental-scale models

So far, each model has been fit to one site at a time, with no information 
(apart from observer effects) shared among sites. When continental-scale data 
is available, ecologists often use correlative models to predict richness 
changes over time from environmental data. Here, we tested three variants of 
this approach. Here, we did *not* include the site-level random effects as 
predictors, meaning that this approach implicitly assumes that two sites 
with identical Bioclim, elevation, and NDVI values should have identical 
richness distributions. As above, we included observer effects & the 
associated uncertainty by running this model [[N]] times (once per MCMC 
sample).

**"Macroecological" model.** The most straightforward way to predict richness is 
to fit a richness model. We chose the `gbm` package for fitting boosted 
regression trees as our example of such a model. Boosted regression trees is a 
common approach in the species distribution modeling literature [refs] and 
works by fitting thousands of small tree-structured models sequentially, with
each tree optimized to reduce the error of its predecessors. For the reasons 
discussed above, this model was optimized using a Gaussian likelihood. [[other 
gbm settings go here.]] Compared to SDMs, this lets us predict richness directly 
instead of using hundreds of noisy species-level estimates, but it also discards 
any potentially-useful information regarding species turnover.

**SDMs.**
* Probably the most common approach?
* These models say that species' occurrence probabilities depend only on the values of
  the environmental variables at the time and place of the transect, not on the history
  of the environment or the history of the birds at that site
* Predict each species' occurrence probabilities individually, then add up the predictions
* [[Something about not using thresholds that throw away information]]
* Taking this approach in the face of uncertainty about observer effects would either 
  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
  times the number of species.
* To make it feasible to fit this large number of models, we used a custom wrapper 
  around the randomForest package. Random forests are flexible models that are built 
  by combining many independent sub-models ("trees") fit to different versions of the training
  data, which made it straightforward to adapt them to our situation: we simply fit one 
  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
  predictions of the resulting forest.
* For the uncertainty, we used Gaussian approximations, based on the means and variances
  of the sums of independent Bernoulli random variables (cite Calabrese et al.).

@@

### Ensembles
* Ensembles. Averaging predictions from multiple models tends to reduce the 
  noise in the estimates, and can thus lead to better predictions. Choosing the
  weights for the ensemble isn't straightforward because our estimates of model 
  error on the training set are biased in favor of models that
  overfit. [[Another option is to spatially cross-validate on the training set]]

### Model evaluation

* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
* When a model was run multiple times (e.g. once per MCMC sample for the observer
  estimates), we used the mean of those runs' point estimates as our final point 
  estimate and we calculated the uncertainty using the law of total variance (i.e.
  the average of the model runs' variance, plus the variance in the point estimates).
* Metrics:
  * (R)MSE 
    * How far, on average, are the models' predictions from the true value?
    * [[RMSE has better units than MSE (species versus squared species), but MSE is
      nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
  * Calibration/bias
    * In general, or in specific situations, is the model consistently predicting
      that richness will be too high or too low?
    * Can also report this value in terms of RMSE or MSE
  * Coverage of 95% confidence intervals
  * Gaussian deviance (simultaneously rewards good point estimates, precision, 
    and coverage)
